Parameter-less Population Pyramid

### 2014-03-18 Parameterless Population Pyramid/main.tex
Minor fixes to P3, added in high level pseudocode

\documentclass{sig-alternate}
%\linespread{2.5}
\usepackage{url}
\usepackage{verbatim}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}

@@

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black box problems is a fundamental goal of evolutionary
computation.  Achieving this goal requires an optimizer to make
assumptions about the nature of the problem it is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.
Many genetic algorithms (GAs) introduce evolutionary parameters which provide
user control of search assumptions.  These parameters can have enormous impact
on search efficiency~\cite{goldberg:1991:gasize} and can have non-obvious optimal
settings which require time consuming tuning to discover~\cite{grefenstette:1986:optimalga}.
When applied to black box problems with non-trivial evaluation functions, the
tuning process can be intractable, forcing users to rely on stock values or attempt
a small number of test values.

@@

The problem of parameter tuning has encouraged work to remove parameters from
user control in an effort to improve out-of-the-box GA quality. In
an early example, a crossover based GA was devised which used
concurrent racing populations which removed all user specified
parameters~\cite{harik:1999:parameterlessga}.
Population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}.
Without the need to set a population size, racing ensures the GA will take no more than
a logarithmic factor increase in evaluations beyond optimal.
While experimental evidence suggests racing is only a constant
factor loss, this method explicitly trades efficiency for applicability.

@@

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many EC techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optima using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to perform crossover,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

@@

The levels represent independent sets of unique solutions, such that
$\forall_{i, j \in P, i \neq j} P_i \cap P_j = \emptyset$.  This relationship
is maintained constructively, such that new solutions can only be added to a level
if they do not exist in any level of the pyramid.  This uniqueness check can be
achieved in constant time by maintaining a hash of all solutions in the pyramid.

@@

Finally, the use of levels allows P3 to perform solution filtering
without throwing away information.  Once found, a local optima is
stored indefinitely, preserving both its linkage and genes for future
crossovers.  By splitting solutions into levels, the linkage
information and the donated genes come from similar solution
qualities.  This prevents noise in gene linkage which only exists in
low quality solutions from effecting cluster formation for high
quality solutions.  It also reduces the chance that optimized
solutions will overwrite newer solutions, preserving
search diversity.

@@

If P3's quality relies solely on starting at random points and
applying FIHC, this comparison algorithm should perform even better
than P3.  Like P3, the FHIC-alone algorithm is parameter-less, makes no
assumptions about gene ordering, and will find the global optima given
a sufficient (potentially exponential) number of evaluations.

\subsection{Linkage Tree Genetic Algorithm}
\label{sec-ltga}
The closest relative to P3 in the existing literature is the Linkage Tree Genetic Algorithm
(LTGA).  There have been a number of different variations proposed to LTGA since
its original publication~\cite{thierens:2010:ltga}, so for comparison we have chosen
to use the variant described in the most recent publication~\cite{thierens:2013:ltgahiff}.
This algorithm is also the current state of the art, outperforming other algorithms
in black box optimization across numerous benchmarks.

@@

\begin{comment}
Since LTGA requires a population size parameter, we thought it most fair to
compare against LTGA with an optimized population size, even though P3
receives no such tuning. To set the population size parameter, we use
the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
size required to meet a specified success criteria.  As P3 will run until the global optima
is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
on its success.  Therefore, using the Rule of Three~\cite{jovanovic:1997:ruleofthree}, we
can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
where $k$ is a number of runs without a failure.  Here we state that a population size
is successful if it performs 100 runs without failing to find the global optimum, meaning
the probability of that population failing to solve problems of the same class is bounded above by
$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
runs are not included during the comparative testing.  Furthermore, when testing on problem
classes, bisection is performed on a different set of instances than testing.
\end{comment}

@@

\section{Test Problems}
All four of our optimization techniques share some interesting properties for
problem equivalence.  All are fitness scale invariant, in that changing
the fitness values without changing the relation between solution fitnesses will not effect search.
\begin{comment}This is true because none of the methods
uses the fitness values directly, only when comparing the relative fitness between
two solutions using boolean operators.  This does not hold true for some optimization
methods, for instance fitness proportional selection.
\end{comment}
None of the methods have a bias toward creating specific gene values, in that
XORing all solutions with a fixed string before evaluation will not effect search speed.
\begin{comment}
This means all of the optimizers
will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
an example, the problem of \emph{maximize the number of ones} in a solution will be treated identically
to the problem of \emph{maximize the number of zeros} in a solution.  Both are also identical
to the problem of minimizing the hamming distance to any arbitrary solution.
\end{comment}
All of the methods are gene order independent, in that the locations of genes
in the genome can be shuffled without effecting search.
\begin{comment}
This means that the location
of genes in the genome can be randomly shuffled without effecting search behavior.
This is not true in many evolutionary systems due to crossover.  For example,
when using one point crossover the distance between genes in the genome effects the
probability they will be kept together through a crossover.
\end{comment}
Combined, these features mean that tests on apparently simple problems can be
indicative of search behavior on a large class of actually complex problems.

In choosing test problems, we decided to use only benchmarks with knowable global
optima.  This allows us to measure how many evaluations each method requires to converge to the
global optima, as opposed to just the highest fitness reached.  This is advantageous
as three of the four algorithms have no hard termination condition.  The remaining
algorithm, LTGA, uses a population parameter which can be increased to slow
convergence while simultaneously increasing the fitness reachable before convergence.
Therefore it is difficult to say how and when to declare any of the techniques
converged unless the global optimum is found.

@@

The classic example for this kind of test is the Deceptive Trap problem.
For this benchmark the genome is broken up into non-overlapping ``traps'' of $k$
genes, with the fitness of each trap given by Equation~\ref{eq-deceptivetrap}.
\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t, &  t<k\\
       k,   &  t = k
     \end{array}
   \right.
  \label{eq-deceptivetrap}
\end{equation}
In this equation $t$ represents the sum of gene values in the trap.  For each trap
there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
We chose to use a trap size of $k=7$.
This problem requires exponential time to solve unless there is some method to effectively
move complete traps between solutions, making this an excellent crossover test problem.

@@

Our final single instance problem is borrowed from the real valued optimization domain.
The Discretized Rastrigin problem uses the standard Rastrigin evaluation function (Equation~\ref{eq-rast}).
\begin{equation}
  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
  \label{eq-rast}
\end{equation}
This function creates a highly multimodal landscape which is completely separable
and has a single global optima.  In order to apply our algorithms to this complex problem,
we discretized the representation using 10 bit gray coded numbers evenly distributed
on the range $[-5.12,5.12)$.

@@

Unlike many theory based algorithms, we believe P3 can achieve bounds on more
complex and interesting landscapes.  Consider this proof sketch on the Deceptive
Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
random individual and optimize it to a local optima using the hill climber.  In
$P_0$ each trap in all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters
will move only complete traps between solutions, and once there is sufficient
diversity in the population there will be a cluster for each individual trap.
When performing a crossover of a single trap there can only be a single alternative
value, meaning if the candidate solution has all 0s the donor must have all 1s, or vice
versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
individual and has a cluster that identifies that trap, all solutions produced by crossing
$P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
all traps we can say that once any $P_i$ has at least one solution with the optimized
version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
The only remaining portion of the proof is to determine how many evaluations are required to reach this point.

>>>>

\documentclass{sig-alternate}
%\linespread{2.5}
\usepackage{url}
\usepackage{verbatim}
\usepackage[noend]{algpseudocode}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}

@@

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black-box problems is a fundamental goal of evolutionary
computation.  Achieving this goal requires an optimizer to make
assumptions about the nature of the problem it is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.
Many genetic algorithms (GAs) introduce evolutionary parameters which provide
user control of search assumptions.  These parameters can have enormous impact
on search efficiency~\cite{goldberg:1991:gasize} and can have non-obvious optimal
settings which require time-consuming tuning to discover~\cite{grefenstette:1986:optimalga}.
When applied to black-box problems with non-trivial evaluation functions, the
tuning process can be intractable, forcing users to rely on stock values or attempt
a small number of test values.

@@

The problem of parameter tuning has encouraged work to remove parameters from
user control in an effort to improve out-of-the-box GA quality. In
an early example, a crossover-based GA was devised which used
concurrent racing populations which removed all user specified
parameters~\cite{harik:1999:parameterlessga}.
Population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}.
Without the need to set a population size, racing ensures the GA will take no more than
a logarithmic factor increase in evaluations beyond optimal.
While experimental evidence suggests racing is only a constant
factor loss, this method explicitly trades efficiency for applicability.

@@

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many GA techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid-like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optimum using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to perform crossover,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

\begin{figure}
  \begin{algorithmic}
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber (Section~\ref{sec-hillclimber})
    \If{solution $\notin$ pyramid}
      \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
      \If{solution fitness has improved}
        \If{solution $\notin$ pyramid}
          \State Add solution to $P_{i+1}$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization.}
  \label{fig:psuedo}
\end{figure}

@@

The levels represent disjoint sets of unique solutions, such that
$\forall_{i, j \in P, i \neq j} P_i \cap P_j = \emptyset$.  This relationship
is maintained constructively, such that new solutions can only be added to a level
if they do not exist in any level of the pyramid.  This uniqueness check can be
achieved in constant time by maintaining a hash of all solutions in the pyramid.

@@

Finally, the use of levels allows P3 to perform solution filtering
without throwing away information.  Once found, a local optimum is
stored indefinitely, preserving both its linkage and genes for future
crossovers.  By splitting solutions into levels, the linkage
information and the donated genes come from similar solution
qualities.  This prevents noise in gene linkage which only exists in
low quality solutions from effecting cluster formation for high
quality solutions.  It also reduces the chance that optimized
solutions will overwrite newer solutions, preserving
search diversity.

@@

If P3's quality relies solely on starting at random points and
applying FIHC, this comparison algorithm should perform even better
than P3.  Like P3, the FHIC-alone algorithm is parameter-less, makes no
assumptions about gene ordering, and will find the global optimum given
a sufficient (potentially exponential) number of evaluations.

\subsection{Linkage Tree Genetic Algorithm}
\label{sec-ltga}
The closest relative to P3 in the existing literature is the Linkage Tree Genetic Algorithm
(LTGA).  There have been a number of different variations proposed to LTGA since
its original publication~\cite{thierens:2010:ltga}, so for comparison we have chosen
to use the variant described in the most recent publication~\cite{thierens:2013:ltgahiff}.
This algorithm is also the current state of the art, outperforming other algorithms
in black-box optimization across numerous benchmarks.

@@

\begin{comment}
Since LTGA requires a population size parameter, we thought it most fair to
compare against LTGA with an optimized population size, even though P3
receives no such tuning. To set the population size parameter, we use
the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
size required to meet a specified success criteria.  As P3 will run until the global optimum
is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
on its success.  Therefore, using the Rule of Three~\cite{jovanovic:1997:ruleofthree}, we
can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
where $k$ is a number of runs without a failure.  Here we state that a population size
is successful if it performs 100 runs without failing to find the global optimum, meaning
the probability of that population failing to solve problems of the same class is bounded above by
$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
runs are not included during the comparative testing.  Furthermore, when testing on problem
classes, bisection is performed on a different set of instances than testing.
\end{comment}

@@

\section{Test Problems}
All four of our optimization techniques share some interesting properties for
problem equivalence.  All are fitness scale invariant, in that changing
the fitness values without changing the relation between solution fitnesses will not affect search.
\begin{comment}This is true because none of the methods
uses the fitness values directly, only when comparing the relative fitness between
two solutions using boolean operators.  This does not hold true for some optimization
methods, for instance fitness proportional selection.
\end{comment}
None of the methods have a bias toward creating specific gene values, in that
XORing all solutions with a fixed string before evaluation will not affect search speed.
\begin{comment}
This means all of the optimizers
will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
an example, the problem of \emph{maximize the number of ones} in a solution will be treated identically
to the problem of \emph{maximize the number of zeros} in a solution.  Both are also identical
to the problem of minimizing the hamming distance to any arbitrary solution.
\end{comment}
All of the methods are gene order independent, in that the locations of genes
in the genome can be shuffled without effecting search.
\begin{comment}
This means that the location
of genes in the genome can be randomly shuffled without effecting search behavior.
This is not true in many evolutionary systems due to crossover.  For example,
when using one point crossover the distance between genes in the genome effects the
probability they will be kept together through a crossover.
\end{comment}
Combined, these features mean that tests on apparently simple problems can be
indicative of search behavior on a large class of actually complex problems.

In choosing test problems, we decided to use only benchmarks with knowable global
optimum.  This allows us to measure how many evaluations each method requires to converge to the
global optimum, as opposed to just the highest fitness reached.  This is advantageous
as three of the four algorithms have no hard termination condition.  The remaining
algorithm, LTGA, uses a population parameter which can be increased to slow
convergence while simultaneously increasing the fitness reachable before convergence.
Therefore it is difficult to say how and when to declare any of the techniques
converged unless the global optimum is found.

@@

The classic example for this kind of test is the Deceptive Trap problem.
For this benchmark the genome is broken up into non-overlapping ``traps'' of $k$
genes, with the fitness of each trap given by Equation~\ref{eq-deceptivetrap}.
\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t, &  t<k\\
       k,   &  t = k
     \end{array}
   \right.
  \label{eq-deceptivetrap}
\end{equation}
In this equation $t$ represents the sum of gene values in the trap.  For each trap
there is one local optimum (all bits set to 0) and one global optimum (all bits set to 1).
We chose to use a trap size of $k=7$.
This problem requires exponential time to solve unless there is some method to effectively
move complete traps between solutions, making this an excellent crossover test problem.

@@

Our final single instance problem is borrowed from the real valued optimization domain.
The Discretized Rastrigin problem uses the standard Rastrigin evaluation function (Equation~\ref{eq-rast}).
\begin{equation}
  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
  \label{eq-rast}
\end{equation}
This function creates a highly multimodal landscape which is completely separable
and has a single global optimum.  In order to apply our algorithms to this complex problem,
we discretized the representation using 10 bit gray coded numbers evenly distributed
on the range $[-5.12,5.12)$.

@@

Unlike many theory based algorithms, we believe P3 can achieve bounds on more
complex and interesting landscapes.  Consider this proof sketch on the Deceptive
Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
random individual and optimize it to a local optimum using the hill climber.  In
$P_0$ each trap in all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters
will move only complete traps between solutions, and once there is sufficient
diversity in the population there will be a cluster for each individual trap.
When performing a crossover of a single trap there can only be a single alternative
value, meaning if the candidate solution has all 0s the donor must have all 1s, or vice
versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
individual and has a cluster that identifies that trap, all solutions produced by crossing
$P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
all traps we can say that once any $P_i$ has at least one solution with the optimized
version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
The only remaining portion of the proof is to determine how many evaluations are required to reach this point.
